---
title: "Power_Plant_Energy"
author: "Kyle Ruhl"
date: "5/1/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is a sample R script assignment for exploring the PowerPlant dataset.

The dataset, from the paper "The environmental footprint of data centers in the United States" by Md Abu Bakar Siddik et al 2021, explores power plants across the lower 48 states + DC and their power type (coal, biomass, solar, etc), the CO2 Emissions associated with producing the electricity, the water and carbon intensity per MWh of power produced, and the water consumption of the plant. 

Before performing the data analysis, for a quick review of renewable vs fossil fuel energy sources, read through this webpage from the US Energy Information Administration: https://www.eia.gov/energyexplained/renewable-sources/ 

--------------------------------------------------------
Lets start by importing and cleaning the data set. After downloading the csv, run the cell block below.  
```{r}
library(tidyverse)
PowerPlants <- read.csv("PowerPlants.csv", header = TRUE)

#Keep variables of interest: 
PowerPlants <- PowerPlants %>% select(c(Plant.state, Plant.name,
                              Latitude, Longitude, Plant.primary.fuel.code,
                              Net.generation..MWh., CO2.e.Emission..tons.,
                              Water.intensity..m3.MWh., Carbon.intensity..Tons.MWh.,
                              Water.Consumption..m3.)) 

#Change variable names:
names(PowerPlants) <- c("Plant.State", "Plant.Name", "Latitude", "Longitude",
                        "Power.Type", "Net.Power.Generated", "CO2.Emissions",
                        "Water.Intensity", "Carbon.Intensity", "Water.Consumption")

#Factor states and power type into levels:
PowerPlants$Plant.State <- factor(PowerPlants$Plant.State)
PowerPlants$Power.Type <- factor(PowerPlants$Power.Type)
```

---------------------------------------------------------
# Task 1: 
Lets explore the breakdown of power type in your home state (or favorite state if you are international!)

1.1: Subset the dataframe so that you are only looking at your chosen state's data. Name this dataframe State1. Output the head of this new dataframe. 

```{r}

```


1.2 Now, lets examine the Power.Type variable in your state. Output a table of the # of power sites using each type of power in your state and save it as table1. 

```{r}
State1$Power.Type = factor(State1$Power.Type) #refactor to remove power sources absent in state


```



1.3 Visualize this table by creating an appropartite plot with correct labels. Next, Describe the plot- what is the most popular power type in your state? Based off your knowledge of your state, is this distribution what you expected? 

```{r}

```

  - ANSWER HERE: 


1.4 Now, lets look at the net power generated by each power type. Use the group_by function to find the sum of the net power generated by each power type in your chosen state, and save this as State1Type. Output the head of this dataframe. 

```{r}

```

1.5 Plot the resulting dataframe State1Type in an approparite visualization to show the sum of the net power generated by each site type in your chosen state. Then, answer the following question: Which site types generated the most power? Did this result match your expectations from 1.3? 
```{r}

```

  - Answer Q here: 
  

1.6 Now we will assess the average power produced by each plant type. Calculate this by taking the net power generated (from 1.5) / by the number of sites in your state (from 1.2). Save this list as the variable By.Plant and plot it using an approparite visualiztion. 

Then, Answer the questions: a) When averaging by the number of sites, did the distribution change for your state? b) Which power types produce the most energy per plant?  c) Compare the energy produced per site for renewable energy sources vs fossil fuel sources. What does this tell us about the number of sites needed for renewable energy to match fossil fuels? 

```{r}

```


Q Answers:
  -a) 
  -b) 
  -c) 


Optional: Consider re-running part 1 with a different state. Particularly, if your original state had 0 or 1 nuclear power plants, consider choosing a state that has 2 or more - good choices are AL,FL,GA,IL,LA,MI,MN,NC,NJ,NY,OH,PA,SC,TN, or VA. If you originally chose one of these, choose a state not on the list and see how it compares. 


#Part 2

Now, lets explore carbon emissions by power type.

2.1: Use group_by to create a new dataframe called Carbon.by.Type that groups the state and power type by the mean of the carbon emissions. Then, create a stacked barplot to visualize, with each state having a bar on one axis, mean of state's carbon emissions on the other axis, and each bar filled in by power type.

For reference, the Carbon.by.Type data frame should look like this
1 AL BIOMASS 1.987808e+04
2 AL COAL 6.967619e+06
3 AL GAS 8.389327e+05
4 AL HYDRO 0.000000e+00
5 AL NUCLEAR 0.000000e+00
6 AL OIL 7.500000e+02
7 AL SOLAR 0.000000e+00
8 AR BIOMASS 1.512933e+04
9 AR COAL 6.573748e+06
... And so on

```{r}

```


2.2 Use group_by to create a stacked barplot of the power plant distribution of the country, where each state is on one axis, the number of power plants on the other, and the fill is energy type. 

```{r}

```

2.3 Based on the stacked barplots from 2.1 and 2.2, do you think carbon emissions from power generation are related to state size/population? Geography? Some other factor? 

  -Answer: 

